---
title: "Impact of Severe Weather Events on Public Health and the Economy in the United States"
author: "JeremiShane"
date: "10/12/2017"
output: 
        html_document: 
                code_folding: show      ## show or hide? default here to show code
                df_print: paged         ## page dataframe output
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```  

## Synopsis


## Environment
```{r system}
s <- Sys.info()
s[3]  ## display sys info
r <- R.Version()
r$platform  ## Display OS and R version
r$version.string
r$nickname
```   

## Include Libraries
```{r includelibraries}
library(plyr)
library(dplyr)
```  

## Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.  

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.

## Data Processing  

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.  

### Data Fetch and Load  

The data come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from:  
https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2  

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.  

- National Weather Service Storm Data Documentation: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf  
- National Climatic Data Center Storm Events FAQ: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf  

```{r readData, cache=TRUE}
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", temp)
df <- read.csv(temp, strip.white=TRUE, na.strings="NA")
unlink(temp)
str(df)
head(df)
```  

### Data Cleanup and Transformation
In this analysis we are attempting to answer the following questions:
(1) Across the United States, which types of events (as indicated in the ð™´ðš…ðšƒðšˆð™¿ð™´ variable) are most harmful with respect to population health?  
(2) Across the United States, which types of events have the greatest economic consequences?  

So we need to isolate the outcome varables that are related to population health.  From visually exploring the data file and documentation these are: $FATALITIES, $INJURIES.  

And we need to isolate the economic outcome variables. According to the documentation the Property damage and crop damage are dollar amounts qualified by a K, M or B to indicate thousands, millions or billions which are indicated in the corresponding EXP variable and are:  $PROPDMG, $PROPDMGEXP, $CROPDMG, $CROPDMGEXP    

According to the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.  For this reason we will only include data from the year 2000 and beyond for our exploration and analysis.

Let's isolate the variables of interest and transform the damage fields into complete dollar amounts that are of the same units of measure.  Then we will look at economic and health impact in very simple terms of the highest quantifications by Event Type.  

The following data frames will be isolated from the original data set:
- dsfe - sum of fatalities by event type
- dsie - sum of injuries by event type
- dpde - sum dollar amount of property damage by event type
- dcde - sum dollar amount of crop damage by event type
- dhealth - sum of fatality points and injury points for health impact score
- deconomic - total economic damage costs (property plus crop) by event type  

```{r variablesofinterest}
## We first assume the association of Fatalities and Injuries as a quantification of impact to population health
dsfe <- subset(df, as.Date(BGN_DATE, "%m/%d/%y") >= "1/1/2000", select=c(FATALITIES, EVTYPE))
dsfe <- dsfe %>% group_by(eventtype=EVTYPE) %>% summarise(fatalities = sum(FATALITIES))
dsfe <- dsfe[order(-dsfe$fatalities), ]

dsie <- subset(df, as.Date(BGN_DATE, "%m/%d/%y") >= "1/1/2000", select=c(INJURIES, EVTYPE))
dsie <- dsie %>% group_by(eventtype=EVTYPE) %>% summarise(injuries = sum(INJURIES))
dsie <- dsie[order(-dsie$injuries), ]

## for our analysis we are going to use a point system for health.  Fatalities are worth 4 points, while injuries are worth 1 point.  

dhealth <- data.frame(eventtype = dsfe$eventtype, fatalitypoints = dsfe$fatalities*4)
di1 <- data.frame(eventtype = dsie$eventtype, injurypoints = dsie$injuries)
dhealth <- merge(dhealth, di1)
dhealth$healthimpactscore <- dhealth$fatalitypoints + dhealth$injurypoints
dhealth <- dhealth[order(-dhealth$healthimpactscore), ]

## Now we look at the variables associated with econimic impact.  It is totally reasonable to think that fatalities and injuries may also have an economic impact.  That would be a good question to answer with further data science analysis. Here we will simply look at health related factors for health impact score separate from financial factors for economic impact score.   

## extract property damage in 1000's
## the variables $PROPDMG, $PROPDMGEXP, $CROPDMG, $CROPDMGEXP  
dpdek <- subset(df, as.Date(BGN_DATE, "%m/%d/%y") >= "1/1/2000" & PROPDMGEXP =="K", select=c(EVTYPE, PROPDMG))
dpdek$PROPDMG <- dpdek$PROPDMG*1000
## extract property damage in 1,000,000's
dpdem <- subset(df, as.Date(BGN_DATE, "%m/%d/%y") >= "1/1/2000" & PROPDMGEXP =="M", select=c(EVTYPE, PROPDMG))
dpdem$PROPDMG <- dpdem$PROPDMG*1000000
## now row bind them together and we have property damage in $
dpde <- rbind(dpdek, dpdem)  

## extract crop damage in 1000's
dcdek <- subset(df, as.Date(BGN_DATE, "%m/%d/%y") >= "1/1/2000" & CROPDMGEXP =="K", select=c(EVTYPE,CROPDMG))
dcdek$CROPDMG <- dcdek$CROPDMG*1000
## extract crop damage in 1,000,000's
dcdem <- subset(df, as.Date(BGN_DATE, "%m/%d/%y") >= "1/1/2000" & CROPDMGEXP =="M", select=c(EVTYPE, CROPDMG))
dcdem$CROPDMG <- dcdem$CROPDMG*1000000
## now row bind them together and we have property damage in $
dcde <- rbind(dcdek, dcdem)

## For economic impact we are going to use $ amount sums of property and crop damage as the measuring stick.
dpde <- dpde %>% group_by(eventtype=EVTYPE) %>% summarise(propertydamage = sum(PROPDMG))
dcde <- dcde %>% group_by(eventtype=EVTYPE) %>% summarise(cropdamage = sum(CROPDMG))
deconomy <- merge(dpde, dcde)
deconomy$economicimpactscore <- deconomy$propertydamage + deconomy$cropdamage
deconomy <- deconomy[order(-deconomy$economicimpactscore), ]

```  

## Results 

This analysis is a very simple one and it could easily be argued there are multiple angles not explored here.  Our results are simply a quantification score for the both health and economic impact.  Fatalities are worth 4x, injury is 1x and summed together for the health impact score.  Our economic impact is simply a sum of $ costs of both property damage and crop damage.  

Below are the top 20 most impactful event types for both health related and economic impact.

```{r theresults}
dhealth[1:20, c(1,4)]  ## top 20 event types for health related impact
deconomy[1:20, c(1,4)]  ## tp 20 event for economy related impact
```


